Concurrent Programming with CXL

### 2020-06-20 book/paper.tex
Latest version

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CVM bytecode.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CVM bytecode.  The first thing we need to
ensure is that there can never be two processes at the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
\glsadd{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

@@

In order to detect violations of progress, and other liveness problems in
algorithms in general, CXL requires that every execution must be
able to reach a state in which all processes have terminated.
Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
the spawned processes never terminate.  We
will model processes in critical sections using the framework in
Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
critical section more than once, but it can also choose to terminate, even
without entering the critical section ever.
(Recall that CXL will try every possible execution.)

@@

Figure~\ref{fig:mutex} shows a high-level state diagram
\index{state diagram}
of the code in
Figure~\ref{fig:cs}.
The circles represent states, while the arrows represent possible state
transitions (high-level steps).  The label in each circle summarizes the state;
the label on an arrow describes the transition.  In this case,
the label contains two numbers: the total number of processes and the number
of processes in the critical section.  A process that is in the critical
section cannot terminate until after leaving the critical section.

@@

You may already have heard of the concept of a \emph{lock}
\index{lock}
\glsadd{lock}
and have realized that
it could be used to implement a critical section.
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
A process that wants to enter the critical section at a time must obtain the
lock first and release it upon exiting the critical section.
Using a lock is a good thought, but how does one implement one?
Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
Initially the lock is not owned, indicated by lock being \texttt{False}.
To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
and then sets it to \texttt{True} to indicate that the lock has been taken.
The process then executes the critical section.  Finally the process
releases the lock by setting it back to \texttt{False}.

@@

To see why, first note the invariant that if process $i$ is in the
critical section, then \texttt{flags[$i$] == True}.
Without loss of generality,
suppose that process~0 sets \texttt{flags[0]} at time $t_0$.
Process 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that \texttt{flags[1] == False}.
Because of the invariant, \texttt{flags[1] == False} implies that
process~1 is not at the critical section at time $t_1$.
Let $t_2$ be the time at which process~0 sets \texttt{flags[0]}
to \texttt{False}.  Process~0 is in the critical section sometime
between $t_1$ and $t_2$.
It is easy to see that process~1 cannot enter the critical section
between $t_1$ and $t_2$, because \texttt{flags[1] == False} at
time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
it would first have to set \texttt{flags[1]} to \texttt{True} and
then wait until \texttt{flags[0] == False}.  But that does not happen
until time $t_2$.

Run the program through CXL.  Unfortunately, it turns out the solution has a
problem: if both try to enter the critical section at the same time, they may
end up waiting for one another indefinitely.  Thus the solution
violates \emph{progress}.

The final na\"{\i}ve solution is based on a variable called \texttt{turn}
that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
enter the critical section, while process $1-i$ has to wait.  When done,
process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
opportunity to enter.
An invariant of this solution is that while process~$i$ is in the critical
section, \texttt{turn} == $i$.
Since \texttt{turn} cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that
processes 0 and 1 alternate entering the critical section.

@@

\section*{Exercises}
\begin{enumerate}
\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
section, mutual is violated, the assertion should fail, and a trace should be reported.
Now plug \texttt{while True: pass;;} before entering the critical section
in Figure~\ref{fig:cs} and run CXL again.  Now mutual exclusion is guaranteed but
progress is violated.  CXL should print a trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports.
\item If we change the loop in Figure~\ref{fig:cs} into an infinite loop,
does the \emph{progress} property guarantee that all processes that are trying
to enter eventually enter?  Why or why not?
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{enumerate}

@@

In 1981, Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
in Figure~\ref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in CXL are atomic.
The process continues to be polite, waiting in a \texttt{while} loop
until either the other process is nowhere near the critical section
(\texttt{flag[1 - self] == False}) or or has given way
(\texttt{turn == self}).
Running the algorithm with CXL shows that it satisfies both mutual
exclusion and progress.

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
\index{state}
\glsadd{state}
at any point of time the CVM is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}
\glsadd{process variable}
for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CVM machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}
\glsadd{step}
Steps in CXL are atomic.

The CVM starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace}
\index{trace}
\glsadd{trace}
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}
if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

@@

One candidate for such an invariant is exclusivity itself.
After all, it certainly holds over the initial state.
And as CXL has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, entering a state that is not exclusive.
So exclusivity is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $C(i) = \mathtt{flags}[1 - i] \land
\mathtt{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
for process~$i$.
Let predicate $I_p(i)$ be the following.
if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $C(i)$ does not hold or process $1-i$ is executing after setting
$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
More formally, $I_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot C(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$I_p(0) \land I_p(1)$.

@@

This anecdote suggests the following.  If you need to do an induction
proof of an algorithm, you have to come up with an inductive invariant.
Before trying to prove the algorithm, you can check that the predicate is
at least invariant by testing it using CXL.  Doing so could potentially
avoid wasting your time on a proof that will not work because the
predicate is not invariant, and therefore not an inductive invariant either.
To finish the story, the author of CXL fixed the Wikipedia page.

>>>>

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CVM bytecode.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CVM bytecode.  The first thing we need to
ensure is that there can never be two processes in the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
\glsadd{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

@@

In order to detect violations of progress, and other liveness problems in
algorithms in general, CXL requires that every execution must be
able to reach a state in which all processes have terminated.
Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
the spawned processes never terminate.  We
will model processes in critical sections using the framework in
Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
critical section more than once, but it can also choose to terminate, even
without entering the critical section ever.
(Recall that CXL will try every possible execution, and so it will evaluate
both choices.)

@@

Figure~\ref{fig:mutex} shows a high-level state diagram
\index{state diagram} of mutual exclusion and progress.
The circles represent states, while the arrows represent possible state
transitions (high-level steps).  The label in each circle summarizes the state;
the label on an arrow describes the transition.  In this case,
the label contains two numbers: the total number of processes and the number
of processes in the critical section.  A process that is in the critical
section cannot terminate until after leaving the critical section.
We will now consider various approaches toward implementing this
specification.

@@

You may already have heard of the concept of a \emph{lock}
\index{lock}
and have realized that
it could be used to implement a critical section.
\glsadd{lock}
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
A process that wants to enter the critical section at a time must obtain the
lock first and release it upon exiting the critical section.
Using a lock is a good thought, but how does one implement one?
Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
Initially the lock is not owned, indicated by lock being \texttt{False}.
To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
and then sets it to \texttt{True} to indicate that the lock has been taken.
The process then executes the critical section.  Finally the process
releases the lock by setting it back to \texttt{False}.

@@

To see why, first note the invariant that if process $i$ is in the
critical section, then \texttt{flags[$i$] == True}.
Without loss of generality,
suppose that process~0 sets \texttt{flags[0]} at time $t_0$.
Process 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that \texttt{flags[1] == False}.
Because of the invariant, \texttt{flags[1] == False} implies that
process~1 is not in the critical section at time $t_1$.
Let $t_2$ be the time at which process~0 sets \texttt{flags[0]}
to \texttt{False}.  Process~0 is in the critical section sometime
between $t_1$ and $t_2$.
It is easy to see that process~1 cannot enter the critical section
between $t_1$ and $t_2$, because \texttt{flags[1] == False} at
time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
it would first have to set \texttt{flags[1]} to \texttt{True} and
then wait until \texttt{flags[0] == False}.  But that does not happen
until time $t_2$.

However, if you run the program through CXL, it turns out the solution
does has a problem: if both try to enter the critical section at the same
time, they may end up waiting for one another indefinitely.  Thus the
solution violates \emph{progress}.

The final na\"{\i}ve solution that we propose
is based on a variable called \texttt{turn}
that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
enter the critical section, while process $1-i$ has to wait.  When done,
process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
opportunity to enter.
An invariant of this solution is that while process~$i$ is in the critical
section, \texttt{turn} == $i$.
Since \texttt{turn} cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that
processes 0 and 1 alternate entering the critical section.

@@

\section*{Exercises}
\begin{enumerate}
\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
section, mutual is violated, the assertion should fail, and a trace should be reported.
Now plug the following code \texttt{while True: pass;;}
just before entering the critical section
in Figure~\ref{fig:cs} and run CXL again.
Mutual exclusion is guaranteed but progress is violated.
CXL should print a short trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{enumerate}

@@

In 1981, Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
in Figure~\ref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in CXL are atomic.
The process continues to be polite, waiting in a \texttt{while} loop
until either the other process is nowhere near the critical section
(\texttt{flag[1 - self] == False}) or has given way
(\texttt{turn == self}).
Running the algorithm with CXL shows that it satisfies both mutual
exclusion and progress.

@@

\glsadd{state}
\glsadd{process variable}
Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
\index{state}
at any point in time the CVM is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}
for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CVM machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}
Steps in CXL are atomic.
\glsadd{step}

\glsadd{trace}
The CVM starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace}
\index{trace}
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}
if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

@@

One candidate for such an invariant is exclusivity itself.
After all, it certainly holds over the initial state.
And as CXL has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, entering a state that is not exclusive.
So exclusivity is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $C(i) = \mathtt{flags}[1 - i] \land
\mathtt{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
for process~$i$.
Let predicate $I_p(i)$ be the following.
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $C(i)$ does not hold or process $1-i$ is executing after setting
$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
More formally, $I_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot C(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$I_p(0) \land I_p(1)$.

@@

This anecdote suggests the following.  If you need to do an induction
proof of an algorithm, you have to come up with an inductive invariant.
Before trying to prove the algorithm, you can check that the predicate is
at least invariant by testing it using CXL.  Doing so could potentially
avoid wasting your time on a proof that will not work because the
predicate is not invariant, and therefore not an inductive invariant either.
(The author fixed the Wikipedia page.)
